Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-26 main.tex
Switch to numbered citations.

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=authoryear]{biblatex}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage[]{todonotes}  % use [disable] to hide notes

@@

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like the natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citep{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \cite{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka clearly showed that steer
torque driven by roll angle feedback stabilizes the capsize mode, by roll
angular rate feedback stabilizes the weave mode, and by steer angular rate
feedback stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle)
eigenmodes are defined in \citep{Sharp1971}.} They also showed how the control
gains must change with respect to vehicle speed for favorable control across
all speeds. This roll motion feedback enables the simplest controller that can
stabilize a single-track vehicle above a minimum speed when one is not
concerned with wobble instabilities. But Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, i.e no human rider was involved.

@@

\appendix
%
\section{Physical Parameters}
\label{sec:physical-parameters}
%
Physical parameters and estimated uncertainties (standard deviation) for the
balance assist bicycle with and without a rider. These are found using the
measurement approach for geometric and inertial parameters for a bicycle
explained in \citet{Moore2012}. The rider inertial parameters are derived from
the 83.5~\si{\kilogram} rider ``Jason'' available in BicycleParameters
1.1.1~\citep{Moore2024}. The variable names match the notation and definitions
from the benchmark bicycle model presented in
\citet{Meijaard2007}.\todo{Uncertainties are based on best guess for
uncertainty in the raw measurements, may be better to remove them as they
aren't so meaningful.}
%
\begin{table}[h]  % have to use "h" so that it stays within the appendix
  \centering
  \caption{Physical parameters of the balance assist bicycle with and without
    the inertia effects of a rigid rider.}
  \begin{tabular}{llrrrr}
    \toprule
    Variable & Unit & Without Rider & Uncertainty & With Rider & Uncertainty \\
    \midrule
    \(c\) & \si{\meter} & 0.04 & 0.01 & 0.04 & 0.01 \\
    \(g\) & \si{\meter\per\second\squared} & 9.80665 & NA & 9.80665 & NA \\
    \(I_{Bxx}\) & \si{\kilogram\meter\squared} & 1.12 & 0.02 & 19.7 & 0.3 \\
    \(I_{Bxz}\) & \si{\kilogram\meter\squared} & 0.05 & 0.03 & -3.5 & 0.2 \\
    \(I_{Byy}\) & \si{\kilogram\meter\squared} & 3.16 & 0.08 & 22.7 & 0.4 \\
    \(I_{Bzz}\) & \si{\kilogram\meter\squared} & 2.12 & 0.03 & 5.0 & 0.1 \\
    \(I_{Fxx}\) & \si{\kilogram\meter\squared} & 0.0995 & 0.0005 & 0.0995 & 0.0005 \\
    \(I_{Fyy}\) & \si{\kilogram\meter\squared} & 0.1902 & 0.0009 & 0.1902 & 0.0009 \\
    \(I_{Hxx}\) & \si{\kilogram\meter\squared} & 0.298 & 0.003 & 0.298 & 0.003 \\
    \(I_{Hxz}\) & \si{\kilogram\meter\squared} & -0.038 & 0.007 & -0.038 & 0.007 \\
    \(I_{Hyy}\) & \si{\kilogram\meter\squared} & 0.257 & 0.006 & 0.257 & 0.006 \\
    \(I_{Hzz}\) & \si{\kilogram\meter\squared} & 0.057 & 0.003 & 0.057 & 0.003 \\
    \(I_{Rxx}\) & \si{\kilogram\meter\squared} & 0.1023 & 0.0006 & 0.1023 & 0.0006 \\
    \(I_{Ryy}\) & \si{\kilogram\meter\squared}& 0.1887 & 0.0007 & 0.1887 & 0.0007 \\
    \(\lambda\) & \si{\radian} & 0.25 & 0.03 & 0.25 & 0.03 \\
    \(m_B\) & \si{\kilogram} & 22.50 & 0.01 & 106.00 & 0.01 \\
    \(m_F\) & \si{\kilogram} & 2.23 & 0.01 & 2.23 & 0.01 \\
    \(m_H\) & \si{\kilogram} & 4.30 & 0.01 & 4.30 & 0.01 \\
    \(m_R\) & \si{\kilogram} & 4.08 & 0.01 & 4.08 & 0.01 \\
    \(r_F\) & \si{\meter} & 0.3523 & 0.0001 & 0.3523 & 0.0001 \\
    \(r_R\) & \si{\meter} & 0.3489 & 0.0001 & 0.3489 & 0.0001 \\
    \(w\) & \si{\meter} & 1.113 & 0.008 & 1.113 & 0.008 \\
    \(x_B\) & \si{\meter} & 0.52 & 0.01 & 0.374 & 0.002 \\
    \(x_H\) & \si{\meter} & 0.92 & 0.02 & 0.92 & 0.02 \\
    \(z_B\) & \si{\meter} & -0.52 & 0.02 & -1.009 & 0.003 \\
    \(z_H\) & \si{\meter} & -0.860 & 0.006 & -0.860 & 0.006 \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=numeric]{biblatex}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage[]{todonotes}  % use [disable] to hide notes

@@

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citet{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \citet{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka clearly showed that steer
torque driven by roll angle feedback stabilizes the capsize mode, by roll
angular rate feedback stabilizes the weave mode, and by steer angular rate
feedback stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle)
eigenmodes are defined in \citep{Sharp1971}.} They also showed how the control
gains must change with respect to vehicle speed for favorable control across
all speeds. This roll motion feedback enables the simplest controller that can
stabilize a single-track vehicle above a minimum speed when one is not
concerned with wobble instabilities. But Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, i.e no human rider was involved.

@@

\appendix
%
\section{Physical Parameters}
\label{sec:physical-parameters}
%
Physical parameters and estimated uncertainties (standard deviation) for the
balance assist bicycle with and without a rider. These are found using the
measurement approach for geometric and inertial parameters for a bicycle
explained in \citet{Moore2012}. The rider inertial parameters are derived from
the 83.5~\si{\kilogram} rider ``Jason'' available in BicycleParameters
1.1.1~\citep{Moore2024}. The variable names match the notation and definitions
from the benchmark bicycle model presented in
\citet{Meijaard2007}.\todo[inline]{Uncertainties are based on best guess for
uncertainty in the raw measurements, may be better to remove them as they
aren't so meaningful.}
%
\begin{table}[h]  % have to use "h" so that it stays within the appendix
  \centering
  \caption{Physical parameters of the balance assist bicycle with and without
    the inertia effects of a rigid rider.}
  \begin{tabular}{llrrrr}
    \toprule
    Variable & Unit & Without Rider & Uncertainty & With Rider & Uncertainty \\
    \midrule
    \(c\) & \si{\meter} & 0.04 & 0.01 & 0.04 & 0.01 \\
    \(g\) & \si{\meter\per\second\squared} & 9.80665 & NA & 9.80665 & NA \\
    \(I_{Bxx}\) & \si{\kilogram\meter\squared} & 1.12 & 0.02 & 19.7 & 0.3 \\
    \(I_{Bxz}\) & \si{\kilogram\meter\squared} & 0.05 & 0.03 & -3.5 & 0.2 \\
    \(I_{Byy}\) & \si{\kilogram\meter\squared} & 3.16 & 0.08 & 22.7 & 0.4 \\
    \(I_{Bzz}\) & \si{\kilogram\meter\squared} & 2.12 & 0.03 & 5.0 & 0.1 \\
    \(I_{Fxx}\) & \si{\kilogram\meter\squared} & 0.0995 & 0.0005 & 0.0995 & 0.0005 \\
    \(I_{Fyy}\) & \si{\kilogram\meter\squared} & 0.1902 & 0.0009 & 0.1902 & 0.0009 \\
    \(I_{Hxx}\) & \si{\kilogram\meter\squared} & 0.298 & 0.003 & 0.298 & 0.003 \\
    \(I_{Hxz}\) & \si{\kilogram\meter\squared} & -0.038 & 0.007 & -0.038 & 0.007 \\
    \(I_{Hyy}\) & \si{\kilogram\meter\squared} & 0.257 & 0.006 & 0.257 & 0.006 \\
    \(I_{Hzz}\) & \si{\kilogram\meter\squared} & 0.057 & 0.003 & 0.057 & 0.003 \\
    \(I_{Rxx}\) & \si{\kilogram\meter\squared} & 0.1023 & 0.0006 & 0.1023 & 0.0006 \\
    \(I_{Ryy}\) & \si{\kilogram\meter\squared}& 0.1887 & 0.0007 & 0.1887 & 0.0007 \\
    \(\lambda\) & \si{\radian} & 0.25 & 0.03 & 0.25 & 0.03 \\
    \(m_B\) & \si{\kilogram} & 22.50 & 0.01 & 106.00 & 0.01 \\
    \(m_F\) & \si{\kilogram} & 2.23 & 0.01 & 2.23 & 0.01 \\
    \(m_H\) & \si{\kilogram} & 4.30 & 0.01 & 4.30 & 0.01 \\
    \(m_R\) & \si{\kilogram} & 4.08 & 0.01 & 4.08 & 0.01 \\
    \(r_F\) & \si{\meter} & 0.3523 & 0.0001 & 0.3523 & 0.0001 \\
    \(r_R\) & \si{\meter} & 0.3489 & 0.0001 & 0.3489 & 0.0001 \\
    \(w\) & \si{\meter} & 1.113 & 0.008 & 1.113 & 0.008 \\
    \(x_B\) & \si{\meter} & 0.52 & 0.01 & 0.374 & 0.002 \\
    \(x_H\) & \si{\meter} & 0.92 & 0.02 & 0.92 & 0.02 \\
    \(z_B\) & \si{\meter} & -0.52 & 0.02 & -1.009 & 0.003 \\
    \(z_H\) & \si{\meter} & -0.860 & 0.006 & -0.860 & 0.006 \\
    \bottomrule
  \end{tabular}
\end{table}
